---
title: "Tutorial to reproduce the co-correspondence analysis of real data sets"
author: "Benjamin Alric"
date: "`r Sys.Date()`"
header-includes:
  - \usepackage{color, fancyvrb}
output:
  rmdformats::readthedown:
    highlight: tango
    number_section: yes
    css: custom.css
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      eval = FALSE,
                      keep.source = FALSE, 
                      tidy = TRUE, 
                      message = FALSE, 
                      tidy.opts = list(width.cutoff = 60), 
                      fig.align = "center")
```

This supplementary material provides data and R scripts to reproduce the real data set analysis of the paper.

# Preliminary steps
To perform the analysis, some packages must be installed and loaded:

```{r, eval = TRUE}
library(SpiecEasi)
library(Matrix)
library(ade4)
library(adegraphics)
library(cocorresp)
source("CoCA_functions_MER-19-0328_R1.R")
```

We also set the seed to ensure reproducibility:

```{r, eval = TRUE}
set.seed(56)
```

# Case study 1: Microbial eukaryotes
Data are loaded:

```{r, eval = TRUE}
load("microeuk_MER-19-0328_R1.rda")
```

## Transformation of raw data
The two tables with read counts of each OTUs are standardized by dividing by the total number of reads in each sample:

```{r, eval = TRUE}
ame <- sweep(microeuk$autotroph, 1, rowSums(microeuk$autotroph), "/")
hme <- sweep(microeuk$heterotroph, 1, rowSums(microeuk$heterotroph), "/")
```

## Symmetric co-correspondence analysis (sCoCA)
Symmetric co-correspondence analysis is firstly performed with all axes:

```{r, eval = TRUE}
sCoCA.ha <- coca(hme ~ ., ame, method = "symmetric", symmetric = TRUE)
```

Permutation test is performed to evaluate the significance of the association between two tables used in sCoCA:

```{r, eval = TRUE, cache = TRUE}
eig.test <- randtest.coca(X = ame, Y = hme, nrepet = 999)
```

The screeplot displays the amount of covariance explained by each axis of the sCoCA:

```{r, eval = TRUE}
screeplot(sCoCA.ha)
```

The first three axes are retained, as suggested by the screeplot.   

The percentage of co-structure explained by the first three axes of the sCoCA can be computed:

```{r, eval = TRUE, cache = TRUE, results = 'markup'}
100 * (sCoCA.ha$lambda[1:3]) / sum(sCoCA.ha$lambda)
```

The Pearson product-moment correlation coefficient between the sample scores can be computed for each axis:

```{r, eval = TRUE}
head(corAxis(sCoCA.ha))
```

We can also compute the variance explained for each table by the sCoCA (with 3 axes):

```{r, eval = TRUE}
sCoCA.ha3 <- coca(hme ~ ., ame, method = "symmetric", symmetric = TRUE, n.axes = 3)
eig.hme <- 100 * (sCoCA.ha3$inertia$total$Y - sCoCA.ha3$inertia$residual$Y) / sCoCA.ha3$inertia$total$Y
eig.ame <- 100 * (sCoCA.ha3$inertia$total$X - sCoCA.ha3$inertia$residual$X) / sCoCA.ha3$inertia$total$X
eig.hme
eig.ame
```

Ordination biplots of sCoCA can be produced (Figure 1 in the paper) using functions of the package `adegraphics`:

Here, the OTU scores ($\mathbf{v}_1^*$ and $\mathbf{\mathbf{v}_2^*}$) are used with sample scores ($\mathbf{u}_1$ and $\mathbf{u}_2$) so that a sample is located at the center of the OTUs it contains. The analysis maximizes the covariance between the two sets of sample scores.

```{r}
g1 <- s.class(sCoCA.ha$scores$species$Y, fac = as.factor(as.character(microeuk$h_id$taxa)), 
              col = c("blue", "red", "green", "turquoise", "orange", "brown", "black", "grey40", "magenta"), 
              psub = list(text = "(a)", position = "topleft"), ellipseSize = 0, starSize = 0, 
              plabel.cex = 0, ppoints.pch = 17, plot = FALSE, key = list(y = 0.01, x = 0.3))
g2 <- s.class(sCoCA.ha$scores$site$Y, 
              fac = factor(microeuk$sample_id$id.sample, levels(microeuk$sample_id$id.sample)[c(1:3, 5, 4)]),
              col = c("red", "dodgerblue3", "forestgreen", "darkorange", "orchid4"), 
              chullSize = 1,ppoints.pch = 5, plabels.optim = TRUE, plabels.cex = "", 
              ppoints.cex = 0.8, plot = FALSE, ellipseSize = 0, starSize = 0, 
              key = list(y = 0.01, x = 0.047))
g3 <- s.class(sCoCA.ha$scores$species$X, fac = as.factor(as.character(microeuk$a_id$taxa)), 
              col = c("blue", "red", "green", "brown", "black", "grey40", "magenta"), 
              psub = list(text = "(b)", position = "topleft"), ellipseSize = 0, starSize = 0, 
              plabel.cex = 0, ppoints.pch = 17, plot = FALSE, key = list(y = 0.01, x = 0.3))
g4 <- s.class(sCoCA.ha$scores$site$X, 
              fac = factor(microeuk$sample_id$id.sample, levels(microeuk$sample_id$id.sample)[c(1:3, 5, 4)]),
              col = c("red", "dodgerblue3", "forestgreen", "darkorange", "orchid4"), 
              chullSize = 1,ppoints.pch = 5, plabels.optim = TRUE, plabels.cex = "", 
              ppoints.cex = 0.8, plot = FALSE, ellipseSize = 0, starSize = 0, 
              key = list(y = 0.01, x = 0.047))

Fig1a <- g2 + g1
Fig1b <- g4 + g3
ADEgS(list(Fig1a, Fig1b))
```

## Infering the co-occurrence network between heterotrophic and autotrophic microbial eukaryotes

We used a novel extension of SPIEC-EASI method (Tipton et al., 2018) of the package `SpiecEasi` (Kurtz et al., 2015) to infer the cross-group co-occurrence network between two tables of heterotrophic and autotrophic microbial eukaryotes.
We used the neighborhood (MB) setting and selected the optimal sparsity parameter based on the Stability Approach to Regularization Selection (StARS).
The StARS variability threshold was set to 0.05 for the co-occurrence network.

```{r}
se.mb.ha <- spiec.easi(list(as.matrix(microeuk$heterotroph), as.matrix(microeuk$autotroph)), 
                       method = 'mb', nlambda = 40, lambda.min.ratio = 1e-4,
                       sel.criterion = 'stars', pulsar.params = list(thresh = 0.05))
```

The association matrix is extracted from SPIEC-EASI analysis:

```{r}
se.network.ha <- as.matrix(symBeta(getOptBeta(se.mb.ha), mode = 'maxabs')) * as.matrix(se.mb.ha$refit$stars)
id.OTU <- c(microeuk$h_id$id.OTU, microeuk$a_id$id.OTU)
colnames(se.network.ha) <- rownames(se.network.ha) <- id.OTU
```

We keep only the associations between heterotrophic and autotrophic microbial eukaryotes in the co-occurrence network:

```{r}
se.ha <- se.network.ha[rownames(se.network.ha) %in% microeuk$h_id$id.OTU, colnames(se.network.ha) %in% microeuk$a_id$id.OTU]
```

Then, the association matrix is transformed from continuous values to discrete ones by replacing negative values by -1 and positive values by 1:

```{r}
se.ha[se.ha > 0] <- 1
se.ha[se.ha < 0] <- -1
```

Finally, the association matrix is reordered according to the OTU scores on the first axis of sCoCA:

```{r}
se.ha.sCoCA <- se.ha[order(sCoCA.ha$scores$species$Y[, 1]), order(sCoCA.ha$scores$species$X[, 1])]
```

The inferred co-occurrence network before and after reordering the position of each OTU according to their scores on the first axis of sCoCA can be produced (Figure 2 in the paper) using functions of the package `adegraphics`:

Just noted that the figures presented here are mirrored from those presented in the paper (in multivariate methods, axes can be flipped without impacting the results and their interpretation).

```{r}
p1 <- table.image(se.ha,
                  xlab = list("\nAutotrophic microbial eukaryotes", cex = 1.1),
                  ylab = list("Heterotrophic microbial eukaryotes\n", cex = 1.1),
                  col = c("magenta", "transparent", "blue", "black"),
                  labelsx = NULL, labelsy = NULL,
                  psub = list(text = "(a)", position = "topleft", cex = 1.3),
                  plot = FALSE)
p2 <- table.image(se.ha.sCoCA,
                  xlab = list("\nAutotrophic microbial eukaryotes", cex = 1.1),
                  ylab = list("Heteretrophic microbial eukaryotes\n", cex = 1.1),
                  col = c("magenta", "transparent", "blue", "black"),
                  labelsx = NULL, labelsy = NULL,
                  psub = list(text = "(b)", position = "topleft", cex = 1.3),
                  plot = FALSE)
p3 <- barchart(microeuk$h_proportion, horizontal = TRUE, 
               xlab = list(label = "Proportion of OTUs", cex = 1.1), 
               box.ratio = 1, box.width = 1, par.settings = list(axis.line = list(col = NA)),
               scales = list(draw = TRUE, tck = c(1, 0), col = 1, y = list(col = "transparent"), cex = 1.5),
               border = "transparent", reference = FALSE, 
               col = c("red", "dodgerblue3","forestgreen", "darkorange", "orchid4"), 
               panel = function(...) {
                 panel.barchart(...)
                 panel.text(0.2, 100, "(d)", cex = 1.5)
               },
               key = list(space = "top", columns = 1, 
                          rectangle = list(col = c("red", "dodgerblue3","forestgreen", "darkorange", "orchid4")),
                          text = list(c("EV", "LC", "MG", "SR", "RSA"), cex = 1.5)))
p4 <- barchart(microeuk$a_proportion, horizontal = FALSE, 
               ylab = list(label = "Proportion of OTUs", cex = 1.1), 
               box.ratio = 1, box.width = 1, par.settings = list(axis.line = list(col = NA)),
               scales = list(draw = TRUE, tck = c(1, 0), col = 1, x = list(col = "transparent"), cex = 1.5),
               border = "transparent", reference = FALSE, 
               col = c("red", "dodgerblue3","forestgreen", "darkorange", "orchid4"),
               panel = function(...) {
                 panel.barchart(...)
                 panel.text(10, 0.9, "(c)", cex = 1.5)
               }
               )

ADEgS(list(p1, p2, p3, p4), positions = rbind(c(0, 0, 0.4, 0.8), 
                                      c(0.4, 0, 0.8, 0.8),
                                      c(0.7, 0, 1, 0.825),
                                      c(0.39, 0.7, 0.75, 1)))
```

# Case study 2: Microalgae-virus system
Data are loaded:

```{r, eval = TRUE}
load("decovir_MER-19-0328_R1.rda")
```

## Transformation of raw data
The two tables with read counts of each OTUs are standardized by dividing by the total number of reads in each sample:

```{r, eval = TRUE}
mdec <- sweep(decovir$microalgae, 1, rowSums(decovir$microalgae), "/")
vdec <- sweep(decovir$virus, 1, rowSums(decovir$virus), "/")
```

## Predictive co-correspondence analysis (pCoCA)
Predictive co-correspondence analysis is firstly performed with all axes:

```{r, eval = TRUE}
pCoCA.vm <- coca(vdec ~ ., data = mdec, method = "predictive", reg.method = "simpls")
```

"Leave-one-out" cross-validation procedure can be used to determine the best number of axes:

```{r, eval = TRUE, cache = TRUE, results = 'hide'}
cross0 <- crossval(vdec, mdec)
```

The number of axes that maximize the CVfit is equal to 3:

```{r, eval = TRUE, cache = TRUE, results = 'markup'}
which.max(cross0$CVfit)
```

A permutation test can also be used to determine the number of significant axes:

```{r, eval = TRUE, cache = TRUE,  results = 'hide'}
pCoCA.perm <- permutest(pCoCA.vm, permutations = 999)
```

It also leads to the conclusion that three dimensions should be kept:

```{r, eval = TRUE, results='markup'}
pCoCA.perm
```

Hence, the pCoCA is performed with only the first two axes:

```{r, eval = TRUE}
pCoCA.vm1 <- coca(vdec ~ ., data = mdec, method = "predictive", reg.method = "simpls", n.axes = 2)
```

Ordination biplots of pCoCA can be produced (Figure 3 in the paper) using functions of the package `adegraphics`:

```{r}
g1 <- s.class(pCoCA.vm1$loadings$X, decovir$m_id$taxa, col = c("dodgerblue3", "darkgreen", "red"), 
              psub = list(text = "(a)", position = "topleft"), ellipseSize = 0, 
              starSize = 0, plabel.cex = 0, ppoints.pch = 17, key = list(x = 0.7, y = 0.8), plot = FALSE)
g2 <- s.label(pCoCA.vm1$scores$site$X, ppoints.pch = 5, plabels.optim = FALSE, 
              plabels = list(box = list(draw = FALSE)),
              plabels.cex = 0.8, ppoints.cex = 0, plot = FALSE)
g3 <- s.class(pCoCA.vm1$scores$species$Y, decovir$v_id$taxa, col = c("dodgerblue3", "darkgreen", "red"),
              psub = list(text = "(b)", position = "topleft"), ellipseSize = 0, 
              starSize = 0, plabel.cex = 0, ppoints.pch = 17, key = list(x = 0.79, y = 0.8), plot = FALSE)

Fig3a <- g2 + g1
Fig3b <- g2 + g3
ADEgS(list(Fig3a, Fig3b))
```

## Infering the co-occurrence network between microalgae and viruses

We applied a novel extension of SPIEC-EASI method (Tipton et al., 2018) of the package `SpiecEasi` (Kurtz et al., 2015) to infer the cross-group co-occurrence network between two tables of microalgae and viruses:

```{r}
se.mb.vm <- spiec.easi(list(as.matrix(decovir$virus), as.matrix(decovir$microalgae)), 
                        method = 'mb', nlambda = 30, lambda.min.ratio = 1e-3,
                        sel.criterion = 'stars', pulsar.params = list(thresh = 0.05))
```

The association matrix is extracted from SPIEC-EASI analysis:

```{r}
se.network.vm <- as.matrix(symBeta(getOptBeta(se.mb.vm), mode='maxabs')) * as.matrix(se.mb.vm$refit$stars)
id.OTU <- c(decovir$v_id$id.OTU, decovir$m_id$id.OTU)
colnames(se.network.vm) <- rownames(se.network.vm) <- id.OTU
```

We keep only the associations between microalgae and viruses in the co-occurrence network:

```{r}
se.vm <- se.network.vm[rownames(se.network.vm) %in% decovir$v_id$id.OTU, colnames(se.network.vm) %in% decovir$m_id$id.OTU]
```

Then, the association matrix is transformed from continuous values to discrete values:

```{r}
se.vm[se.vm > 0] <- 1
se.vm[se.vm < 0] <- -1
```

Finally, the association matrix is reordered according to the OTU scores on the first axis of pCoCA:

```{r}
u1p <- pCoCA.vm1$scores$species$Y # virus
u2.loadings <- pCoCA.vm1$loadings$X # microalgae
se.vm.pCoCA <- se.vm[order(u1p[, 1]), order(u2.loadings[, 1])]
```

The inferred co-occurrence network before and after reordering the position of each OTU according to their scores on the first axis of pCoCA can be produced (Figure 4 in the paper) using functions of the package `adegraphics`:

Just noted that the figures presented here are mirrored from those presented in the paper.

```{r}
p1 <- table.image(se.vm,
                  xlab = list("\nMicroalgae", cex = 1.1),
                  ylab = list("Viruses\n", cex = 1.1),
                  col = c("magenta", "transparent", "blue", "black"),
                  labelsx = NULL, labelsy = NULL,
                  psub = list(text = "(a)", position = "topleft", cex = 1.3),
                  plot = FALSE)
p2 <- table.image(se.vm.pCoCA,
                  xlab = list("\nMicroalgae", cex = 1.1),
                  ylab = list("Viruses\n", cex = 1.1),
                  col = c("magenta", "transparent", "blue", "black"),
                  labelsx = NULL, labelsy = NULL,
                  psub = list(text = "(b)", position = "topleft", cex = 1.3),
                  plot = FALSE)
p3 <- barchart(decovir$v_proportion, horizontal = TRUE, 
               xlab = list(label = "Proportion of OTUs", cex = 1.1), 
               box.ratio = 1, box.width = 1, par.settings = list(axis.line = list(col = NA)),
               scales = list(draw = TRUE, tck = c(1, 0), col = 1, y = list(col = "transparent"), cex = 1.5),
               border = "transparent", reference = FALSE, 
               col = c("red", "dodgerblue3", "forestgreen", "darkorange"), 
               panel = function(...) {
                 panel.barchart(...)
                 panel.text(0.1, 96, "(d)", cex = 1.5)
               },
               key = list(space = "top", columns = 1, 
                          rectangle = list(col = c("red", "dodgerblue3", "forestgreen","darkorange")),
                          text = list(c("LA", "LB", "SA", "MA"), cex = 1.5)))
p4 <- barchart(decovir$m_proportion, horizontal = FALSE, 
               ylab = list(label = "Proportion of OTUs", cex = 1.1), 
               box.ratio = 1, box.width = 1, par.settings = list(axis.line = list(col = NA)),
               scales = list(draw = TRUE, tck = c(1, 0), col = 1, x = list(col = "transparent"), cex = 1.5),
               border = "transparent", reference = FALSE, 
               col = c("red", "dodgerblue3", "forestgreen", "darkorange"),
               panel = function(...) {
                 panel.barchart(...)
                 panel.text(7, 0.94, "(c)", cex = 1.5)
               }
               )

ADEgS(list(p1, p2, p3, p4), positions = rbind(c(0, 0, 0.4, 0.8), 
                                      c(0.4, 0, 0.8, 0.8),
                                      c(0.7, 0, 1, 0.825),
                                      c(0.39, 0.7, 0.75, 1)))
```

# Note about the CoCA and the scale invariance principle

Scale invariance is a condition that any statistical method applied to compositional data should fulfill. The scale invariance principle states that compositional data carry only relative information, so that any change of the scale of the original data has no effect. So if the original data are multiplied by any scale fator C then the compositional data remain the same after the closure. Closure corresponds to the operation of dividing out a set of data by its total to obtain the compositional values, while are proportions adding up to 1. For example, we should obtain exactly the same OTU scores in CoCA if we analyzed percentages (of 100) as we would if we analyzed proportions (of 1).

Here, CoCA computed on proportion data correspond to analyses previously made.

Case study 1:

```{r, eval = TRUE}
sCoCA.ha.prop <- sCoCA.ha
```

Case study 2:

```{r, eval = TRUE}
pCoCA.vm.prop <- pCoCA.vm1
```

Then, percentage are determined.

Case study 1:

```{r, eval = TRUE}
ame.per <- sweep(microeuk$autotroph, 1, rowSums(microeuk$autotroph), "/") * 100
hme.per <- sweep(microeuk$heterotroph, 1, rowSums(microeuk$heterotroph), "/") * 100
```

Case study 2:

```{r, eval = TRUE}
mdec.per <- sweep(decovir$microalgae, 1, rowSums(decovir$microalgae), "/") * 100
vdec.per <- sweep(decovir$virus, 1, rowSums(decovir$virus), "/") * 100
```

CoCA were computed from percentage data.

Case study 1:

```{r, eval = TRUE}
sCoCA.ha.per <- coca(hme.per ~ ., ame.per, method = "symmetric", symmetric = TRUE)
```

Case study 2:

```{r, eval = TRUE}
pCoCA.vm.per <- coca(vdec.per ~ ., mdec.per, method = "predictive", reg.method = "simpls", n.axes = 2)
```

Results are then compared:

```{r, width = 12, height = 12}
par(mfcol = c(2, 2),
    mar = c(4, 4, 1, 1),
    oma = c(1, 2, 1, 1))

plot(sCoCA.ha.prop$scores$species$X[, 1], sCoCA.ha.per$scores$species$X[, 1], yaxt = "n", pch = 19,
     xlab = "autotroph scores (proportion)",
     ylab = "autotroph scores (percentage)",
     ylim = c(-1.0, 1.5), xlim = c(-1.0, 1.5), cex = 0.5)
reg <- lm(sCoCA.ha.prop$loadings$X[, 1] ~ sCoCA.ha.per$loadings$X[, 1])
abline(reg, col = "red")
axis(side = 2, las = 1)
text(-0.975, 1.5, "(a)")
mtext(text = "symmetric CoCA", side = 3, line = 0.5)

plot(sCoCA.ha.prop$scores$species$Y[, 1], sCoCA.ha.per$scores$species$Y[, 1], yaxt = "n", pch = 19,
     xlab = "heterotroph scores (proportion)",
     ylab = "heterotroph scores (percentage)",
     ylim = c(-1.0, 1.5), xlim = c(-1.0, 1.5), cex = 0.5)
reg <- lm(sCoCA.ha.prop$scores$species$Y[, 1] ~ sCoCA.ha.per$scores$species$Y[, 1])
abline(reg, col = "red")
axis(side = 2, las = 1)
text(-0.975, 1.5, "(b)")

plot(pCoCA.vm.prop$loadings$X[, 1], pCoCA.vm.per$loadings$X[, 1], yaxt = "n", pch = 19,
     xlab = "microalgae scores (proportion)",
     ylab = "microalgae scores (percentage)",
     ylim = c(-1.0, 1.5), xlim = c(-1.0, 1.5), cex = 0.5)
reg <- lm(pCoCA.vm.prop$loadings$X[, 1] ~ pCoCA.vm.per$loadings$X[, 1])
abline(reg, col = "red")
axis(side = 2, las = 1)
text(-0.975, 1.5, "(c)")
mtext(text = "predictive CoCA", side = 3, line = 0.5)

plot(pCoCA.vm.prop$scores$species$Y[, 1], pCoCA.vm.per$scores$species$Y[, 1], yaxt = "n", pch = 19,
     xlab = "virus scores (proportion)",
     ylab = "virus scores (percentage)",
     ylim = c(-1.0, 1.5), xlim = c(-1.0, 1.5), cex = 0.5)
reg <- lm(pCoCA.vm.prop$scores$species$Y[, 1] ~ pCoCA.vm.per$scores$species$Y[, 1])
abline(reg, col = "red")
axis(side = 2, las = 1)
text(-0.975, 1.5, "(d)")
```

It is clear that percentage or proportion does not affect OTU scores computed in pCoCA as well as in sCoCA, which confirms that CoCA statisfies scale invariance.  

# Version of packages used to build this document

```{r}
sessionInfo()
```

# References
Kurtz Z. D., Müller C. L., Miraldi E. R., Littman D. R., Blaser M. J., & Bonneau R.A. (2015). Sparse and compositionally robust inference of microbial ecological networks. PLoS Computational Biology, 11, e1004226.

Tipton L., Müller C. L., Kurtz Z. D., Huang L., Kleerup E., Morris A., Bonneau R., & Ghedin E. (2018). Fungi stabilize connectivity in the lung and skin microbial ecosystems. Microbiome, 6, 12.